Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-05 intro.tex
tweaks to sec 2 and 3

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where the edit state is a  program structure, rather than text. % Eliminating text eliminates the possibility of syntax errors.

@@

The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics. For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}. Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design. The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed. As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics.
%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.

@@

This paper develops a principled solution to this problem. We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}, respectively. Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because we follow Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g. a function argument), only actions consistent with that type are permitted.

Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner. For example, the programmer might need to construct the outer function application form before identifying the intended function. To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type. This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}. In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remains under construction. %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}. %Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)

@@

The remainder of this paper is organized as follows:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item We begin in Section
    \ref{sec:example} with two example edit sequences to develop the reader's intuitions.  
\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Section \ref{sec:hazel}.
\item Hazelnut is a {foundational} calculus, i.e. a calculus that language and editor designers are expected to extend with higher level constructs. We show how Hazelnut's rich mechanized metatheory guides the development of an extension of Hazelnut with simple sum types in Section \ref{sec:extending}. 
\item In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.

@@

\item In Section \ref{sec:rw}, we summarize related work. In particular, much of the technical machinery needed to handle type holes coincides with that developed in the study of gradual type systems. Similarly, one interpretation of expression holes is as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work. %This suggests a principled design for an ``edit and resume'' feature.

\item We conclude in Section \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
\end{itemize} 
The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.

>>>>

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where every edit state is a program structure. % Eliminating text eliminates the possibility of syntax errors.

@@

The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics. For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}. Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design. The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed. As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful. 
% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.

@@

This paper develops a principled solution to this problem. We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}. Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g. a function argument), only actions consistent with that type are permitted.

Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner. For example, the programmer would often need to construct the outer function application form before identifying the intended function. To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type. This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}. In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction. %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}. %Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)

@@

The remainder of this paper is organized as follows:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item We begin in Sec. 
    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.  
\item We then give a detailed overview of Hazelnut's semantics and metatheory, which have been mechanized using the Agda proof assistant, in Sec.  \ref{sec:hazel}.
\item Hazelnut is a {foundational} calculus, i.e. a calculus that language and editor designers are expected to extend with higher level constructs. We extend Hazelnut with simple sum types in Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich mechanized metatheory guides such a development. 
\item In Sec.  \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.

@@

\item In Sec.  \ref{sec:rw}, we summarize related work. In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems. Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work. %This suggests a principled design for an ``edit and resume'' feature.

\item We conclude in Sec.  \ref{sec:future} by summarizing our vision of a principled science of structure editor design rooted in type theory, and suggest a number of future directions.
\end{itemize} 
The supplemental material includes 1) a typeset listing of Hazelnut's rules in definitional order; 2) the formalization of Hazelnut in Agda; and 3) our reference implementation of Hazelnut, both in source form and pre-compiled to JavaScript for use in a web browser.
